---
title: "Mt Robson Waterbody Map"
author: "John Phelan"
date: "2026-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE
)

library(tidyverse)
library(sf)
library(dplyr)
library(leaflet)
library(leaflet.extras)
library(bcdata)

lan_root = "//SFP.IDIR.BCGOV/S140/S40203/WFC AEB/General/2 SCIENCE - Invasives/"

wbs<-readRDS(paste0(lan_root,"AIS_R_Projects/LargeDataFiles/named_lakes_and_rivers.rds"))

named_wbs = c("Portal Lake", "Yellowhead Lake", "Moose Lake", "Kinney Lake", "Witney Lake",
                           "Fraser River", "Yellowhead River", "Moose River", "Robson River",
              "Yellowhead Creek", "Berg Lake")

wbsFiltered<- wbs |> 
  filter( waterbody %in% named_wbs)

rm(wbs)
#setdiff(named_wbs, unique(wbsFiltered$waterbody))

my.external.data.folder = "Z:/2 SCIENCE - Invasives/AIS_R_Projects/CMadsen_Wdrive/Projects/ZQMussels/data/"
stations = sf::read_sf(paste0(my.external.data.folder,"inspection_stations.gpkg"))

robson_loc <- stations |>
  dplyr::filter(station_name == "Mt. Robson")

# we are missing some - lets find them
# st_test = bcdc_query_geodata('92344413-8035-4c08-b996-65a9b3f62fca') |>
#   filter(GNIS_NAME == "Fraser River") |>
#   collect() |>
#   sf::st_zm() |>
#   dplyr::group_by(BLUE_LINE_KEY) |>
#   dplyr::summarise() |>
#   sf::st_buffer(dist = 10) |>
#   dplyr::mutate(type = 'stream') |>
#   st_transform(4326)
# 
# leaflet() %>%
#   addProviderTiles(providers$CartoDB.Positron) %>%
# 
#   # Sampled 2025 TRUE
#   addPolygons(
#     data = st_test,
#     color = "blue",
#     fillColor = "blue",
#     fillOpacity = 0.9,
#     stroke = TRUE,
#     weight = 1,
#     popup = ~paste0(
#       "<b>BLK:</b> ", BLUE_LINE_KEY, "<br>"
#     ),
#     group = "Sampled 2025: TRUE"
#   )



robson_river = bcdc_query_geodata('92344413-8035-4c08-b996-65a9b3f62fca') |>
  filter(BLUE_LINE_KEY == 356364407) |>
  collect() |>
  dplyr::group_by(BLUE_LINE_KEY) |>
  dplyr::summarise() |>
  sf::st_buffer(dist = 10) |>
  dplyr::mutate(`Waterbody Name` = "Robson River",
                `Watershed Name` = NA,
                wb_type = "river") |>
  sf::st_transform(4326)

wbsFiltered <- wbsFiltered |> 
  bind_rows(robson_river)


fraser_river = bcdc_query_geodata('92344413-8035-4c08-b996-65a9b3f62fca') |>
  filter(BLUE_LINE_KEY == 356364114) |>
  collect() |>
  dplyr::group_by(BLUE_LINE_KEY) |>
  dplyr::summarise() |>
  sf::st_buffer(dist = 10) |>
  dplyr::mutate(`Waterbody Name` = "Fraser River",
                `Watershed Name` = NA,
                wb_type = "river") |>
  sf::st_transform(4326)

wbsFiltered <- wbsFiltered |> 
  bind_rows(robson_river, fraser_river)


# ## Fraser is broken - lets get a fresh one
# fraser_river = bcdc_query_geodata('freshwater-atlas-rivers') |>
#     filter(!is.na(GNIS_NAME_1)) |>
#     filter(GNIS_NAME_1 == "Fraser River") |> 
#     collect() |>
#     dplyr::select(waterbody = GNIS_NAME_1, watershed = WATERSHED_GROUP_ID, BLUE_LINE_KEY, FWA_WATERSHED_CODE) |>
#     dplyr::group_by(waterbody, watershed, BLUE_LINE_KEY, FWA_WATERSHED_CODE) |>
#     dplyr::summarise(.groups = 'drop') |>
#     dplyr::mutate(wb_type = "river") |> 
#   st_transform(4326)
# 
# 
# # drop the old Fraser river
# wbsFiltered <- wbsFiltered |> 
#   filter(waterbody != "Fraser River")
# 
# 
# wbsFiltered <- wbsFiltered |> 
#   bind_rows(fraser_river)


#unique(wbsFiltered$waterbody)

## Cannot find yellowhead river

wbs_all = wbsFiltered

library(sf)
library(dplyr)

# Make sure both layers use a projected CRS in metres
wbs_proj <- st_transform(wbs_all, 3005)      # BC Albers
robson_proj <- st_transform(robson_loc, 3005)

# 200 km buffer around Mt. Robson
robson_buffer <- st_buffer(robson_proj, dist = 200000)

# Features within 200 km
wbsFiltered <- wbs_proj |>
  st_intersection(robson_buffer)

```




```{r parks_protected_areas_across_BC, fig.width = 9, fig.height = 6}


if(!file.exists("data/parks_prot_areas_bcdata_layer.rds")){
  parks = bcdc_query_geodata('terrestrial-protected-areas-representation-by-ecosection-parc-') |>
    collect()
  # Note: variable PROTECTED_AREA_TYPE has levels PROVINCIAL PARK (1038), NGO CONSERVANCY (603), etc.

  saveRDS(parks, file = "data/parks_prot_areas_bcdata_layer.rds")
} else {
  parks = readRDS("data/parks_prot_areas_bcdata_layer.rds")
}

# parks |> 
#   sf::st_drop_geometry() |> 
#   dplyr::count(PROTECTED_AREA_TYPE, sort = T) |> 
#   knitr::kable()


parksFiltered <- parks |>
  filter(lengths(st_intersects(geometry, robson_buffer)) > 0) |> 
  filter(PROTECTED_AREA_TYPE == "PROVINCIAL PARK")


```

```{r map_fixes}

# Ensure WGS84 for Leaflet
map_data <- st_transform(wbsFiltered, 4326)

map_data <- map_data |>
  mutate(
    popup = paste0("<b>", waterbody, "</b>")
)

pal <- colorFactor(
  c("orange", "steelblue"),
  domain = map_data$wb_type
)

parksFiltered<- st_transform(parksFiltered, 4326)
```


```{r sus_species}

# Susceptible fish species
species_names <- c(
  "Bull Trout", "Sockeye Salmon", "Cutthroat Trout", "Coho Salmon",
  "Rainbow Trout", "Chinook Salmon", "Mountain Whitefish",
  "Atlantic Salmon", "Brown Trout", "Brook Trout",
  "Westslope Cutthroat Trout", "Steelhead",
  "Lake Whitefish", "Round Whitefish", "Pygmy Whitefish"
)

# Read occurrence records
species_occs <- lapply(species_names, function(sp) {
  sf::read_sf(
    here::here(
      "..",
      "AIS_WhirlingDiseaseMonitoring",
      "data",
      paste0(gsub(" ", "_", sp), "_occurrences.gpkg")
    )
  )
})

names(species_occs) <- gsub(" ", "_", species_names)

fish_occs <- dplyr::bind_rows(species_occs, .id = "species")

# Match CRS to your waterbodies
fish_occs <- st_transform(fish_occs, st_crs(map_data))


map_data <- map_data |>
  st_make_valid()

fish_occs <- fish_occs |>
  st_make_valid()

wbs_spp <- st_join(map_data, fish_occs)

# One record per waterbody
wb_species <- wbs_spp |>
  group_by(waterbody) |>
  summarise(
    sus_spp = paste(sort(unique(species[!is.na(species)])), collapse = ", "),
    .groups = "drop"
  ) |>
  mutate(
    sus_spp = ifelse(sus_spp == "", NA_character_, sus_spp)
  ) |>
  st_drop_geometry()

# Join species back onto map_data
map_data <- map_data |>
  left_join(wb_species, by = "waterbody")

```



# {.tabset}


 



## Waterbody Map


```{r map_locations, fig.width = 8, fig.height=10}
leaflet(map_data) |>
  addProviderTiles(providers$CartoDB.Positron) |>

  # Waterbodies
  addPolygons(
    fillColor = ~pal(wb_type),
    color = ~pal(wb_type),
    weight = 2,
    fillOpacity = 1,
    label = ~waterbody,
    group = "Waterbodies",
    popup = ~paste0(
      "<b>", waterbody, "</b><br>",
      "<b>Type:</b> ", wb_type, "<br>",
      "<b>Susceptible species:</b> ",
      ifelse(is.na(sus_spp), "None recorded", sus_spp)
    )
  ) |>

  # Mt. Robson location
  addCircleMarkers(
    data = robson_loc,
    radius = 8,
    color = "black",
    fillColor = "lightgrey",
    fillOpacity = 1,
    stroke = TRUE,
    weight = 2,
    popup = ~station_name,
    label = ~station_name,
    group = "Mt. Robson"
  ) |>

  # Protected Areas
  addPolygons(
    data = parksFiltered,
    fillColor = "darkgreen",
    color = "darkgreen",
    weight = 2,
    fillOpacity = 0.3,
    label = ~PROTECTED_AREA_TYPE,
    popup = ~PROTECTED_AREA_TYPE,
    group = "Provincial Parks"
  ) |>

  # Legend
  addLegend(
    pal = pal,
    values = map_data$wb_type,
    title = "Waterbody type"
  ) |>

  # Layer control
  addLayersControl(
    overlayGroups = c("Provincial Parks", "Mt. Robson"),
    options = layersControlOptions(collapsed = FALSE)
  )

## make kable of waterbodytable
map_data |>
  st_drop_geometry() |>
  dplyr::select(waterbody, wb_type, sus_spp) |>
  knitr::kable(
    caption = "Waterbody details",
    col.names = c("Waterbody", "Type", "Susceptible Species")
  )
  

```


The waterbodies listed are:

 - Portal Lake
 - Yellowhead Lake
 - Moose Lake
 - Kinney Lake - Witney Lake
 - Fraser River
 - Yellowhead River
 - Moose River
 - Robson River
 - Yellowhead Creek
 
The susceptible salmonids included are:

- Bull Trout
- Sockeye Salmon
- Cutthroat Trout
- Coho Salmon
- Rainbow Trout
- Chinook Salmon
- Mountain Whitefish
- Atlantic Salmon
- Brown Trout
- Brook Trout
- Westslope Cutthroat Trout
- Steelhead
- Lake Whitefish
- Round Whitefish
- Pygmy Whitefish

## Watercraft Inspection Data

```{r get_waterbody_use, fig.height= 6, fig.width=8, message=FALSE, warning=FALSE, results='hide'}


source(here::here("02_IMDP_Figure_Generator/utils/initialise_plots.R"))

my.year = 2026



source(here::here("02_IMDP_Figure_Generator", "utils", "load_figure_data.R"))



dat = dat |> 
  mutate(Start_Time = as.POSIXct(
    Start_Time * 86400,
    origin = "1899-12-30",
    tz = "UTC"
  )) |> 
  mutate(End_Time = as.POSIXct(
    End_Time * 86400,
    origin = "1899-12-30",
    tz = "UTC"
  ))

## rename these to be Douglas
dat <- dat |>
  mutate(
    Station = if_else(
      (
        str_detect(Shift_Start_Comment, "DFO") |
          str_detect(Shift_End_Comment, "DFO")
      ) &
        as_date(TimeOfInspection) %in% as.Date(c(
          "2025-07-04", "2025-07-05", "2025-07-06",
          "2025-07-18", "2025-07-19",
          "2025-06-27", "2025-06-28",
          "2025-08-28", "2025-08-29", "2025-08-30"
        )) &
        Station != "Pacific",
      "Douglas Crossing",
      Station
    )
  )

dat_all = dat_all |> 
  mutate(
    Station = if_else(
      (
        str_detect(Shift_Start_Comment, "DFO") |
          str_detect(Shift_End_Comment, "DFO")
      ) &
        as_date(TimeOfInspection) %in% as.Date(c(
          "2025-07-04", "2025-07-05", "2025-07-06",
          "2025-07-18", "2025-07-19",
          "2025-06-27", "2025-06-28",
          "2025-08-28", "2025-08-29", "2025-08-30"
        )) &
        Station != "Pacific",
      "Douglas Crossing",
      Station
    )
  )

source(here::here("02_IMDP_Figure_Generator", "utils", "station_manipulation.R"))


### Okay lets fix some names

dat = dat |> 
  mutate(Station = case_when(
    Station == "Lower Mainland Roving - Douglas"      ~ "Douglas Crossing",
    Station == "Lower Mainland - Douglas"      ~ "Douglas Crossing",
    Station == "Lower Mainland - Huntington"      ~ "Sumas (Huntington)",
    Station == "Lower Mainland - Scheduled Inspection"      ~ "Lower Mainland",
    Station == "Lower Mainland Roving - Outreach"      ~ "Lower Mainland",
    Station == "Lower Mainland Roving - Huntington"      ~ "Sumas (Huntington)",
    TRUE ~ Station
  ))

dat_all = dat_all |> 
  mutate(Station = case_when(
    Station == "Lower Mainland Roving - Douglas"      ~ "Douglas Crossing",
    Station == "Lower Mainland - Douglas"      ~ "Douglas Crossing",
    Station == "Lower Mainland - Huntington"      ~ "Sumas (Huntington)",
    Station == "Lower Mainland - Scheduled Inspection"      ~ "Lower Mainland",
    Station == "Lower Mainland Roving - Outreach"      ~ "Lower Mainland",
    Station == "Lower Mainland Roving - Huntington"      ~ "Sumas (Huntington)",
    TRUE ~ Station
  ))

# source(here::here("02_IMDP_Figure_Generator", "utils", "dat_to_stations.R"))

stations_for_maps <- stations_for_maps |>
  dplyr::filter(station_name != "Radium")
station_labels = station_labels |> 
  dplyr::filter(station_name != "Radium")



source(here::here("02_IMDP_Figure_Generator", "utils", "get_basemaps.R"))

library(stringr)


pattern <- str_c(map_data$waterbody, collapse = "|")

pattern <- str_to_lower(pattern)

dat_3 <- dat_all |>
  filter(
    str_to_lower(Destination_Waterbody_1_Name) %in%
      str_to_lower(map_data$waterbody)
  )

dat_3 <- dat_3[!is.na(dat_3$Destination_Waterbody_1_Name),]



ggplot(dat_3, aes(x = Year)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~ Destination_Waterbody_1_Name, scales = "free_y") +
  scale_x_continuous(breaks = 2015:2026) +
  labs(
    title = "Watercraft Inspections by Year and Destination Waterbody",
    x = "Year",
    y = "Number of Inspections"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  )

```

## Angler Survey Results

```{r anglerSurvey, fig.width= 8, fig.height=6}

# Read occurrence records
fishing_survey <- readRDS(
    here::here(
      "..",
      "AIS_WhirlingDiseaseMonitoring",
      "data",
      "fishing_days_by_waterbody_and_watershed_w_BLK.rds"
    )
  )


# Join fishing_days onto map_data, dropping fishing_survey's geometry first
map_data_joined <- map_data |>
  left_join(
    fishing_survey |>
      st_drop_geometry() |>
      select(waterbody, watershed, fishing_days),
    by = c("waterbody", "watershed")
  )

# Colour palette based on fishing_days
pal <- colorNumeric(
  palette = c(
    "#08306B",  # dark blue
    "#2171B5",
    "#6BAED6",
    "#FDB863",
    "#E66101"   # strong orange
  ),
  domain = map_data_joined$fishing_days,
  na.color = "grey80"
)

# leaflet(map_data_joined) |>
#   addProviderTiles(providers$CartoDB.Positron) |>
#   addPolygons(
#     fillColor = ~pal(fishing_days),
#     fillOpacity = 0.8,
#     color = "grey",
#     weight = 1,
#     label = ~paste0(waterbody, ": ", fishing_days, " fishing days"),
#     highlightOptions = highlightOptions(
#       weight = 2,
#       color = "black",
#       bringToFront = TRUE
#     )
#   ) |>
#   addCircleMarkers(
#     data = robson_loc,
#     radius = 8,
#     color = "black",
#     fillColor = "lightgrey",
#     fillOpacity = 1,
#     stroke = TRUE,
#     weight = 2,
#     popup = ~station_name,
#     label = ~station_name,
#     group = "Mt. Robson"
#   ) |>
#   addLegend(
#     pal = pal,
#     values = ~fishing_days,
#     title = "Fishing Days",
#     position = "bottomright"
#   ) 

DT::datatable(
  map_data_joined |>
    st_drop_geometry() |>
    select(waterbody, watershed, fishing_days) |>
    distinct() |>
    arrange(desc(fishing_days)),
  colnames = c(
    "Waterbody",
    "Watershed",
    "Fishing Days"
  ),
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 15
  )
)

```

